---
title: "Project Title"
subtitle: "INFO 526 - Final Project"
author: 
  - name: "RoboCops - Peter, Sachin, Arun, Surajit, Christian, Valerie"
    affiliations:
      - name: "School of Information, University of Arizona"
description: "This project aims to comprehensively analyze fatal police shootings in the United States, delving into the underlying factors and motivations driving these tragic events."
format:
   html:
    code-tools: true
    code-overflow: wrap
    embed-resources: true
editor: visual
execute:
  warning: false
  echo: false
---

```{r}
#| label: load-packages
#| include: false

# Load packages here
pacman::p_load(dplyr,
               ggplot2)

```

```{r}
#| label: load-dataset
#| message: false

# Load in the datasets
fatality <- read.csv("data/PoliceKillingsUS.csv", na.strings = "")
median_income <- read.csv("data/MedianHouseholdIncome2015.csv")
poverty_perc <- read.csv("data/PercentagePeopleBelowPovertyLevel.csv")
hs_perc <- read.csv("data/PercentOver25CompletedHighSchool.csv")
race_perc <- read.csv("data/ShareRaceByCity.csv")
city_pop <- read.csv("data/us-cities-top-1k-multi-year.csv")
state_pop <- read.csv("data/Population_Estimate_data_Statewise_2010-2023.csv")
```

## Abstract

The objective of this project is to analyze and contextualize police shootings, aiming to understand the underlying reasons and factors contributing to these unfortunate events. Acknowledging the necessity of police interventions for public safety, there has been a concerning uptick of documented unjustified shootings in recent years. We seek to expand the analysis beyond individual cases to explore socioeconomic factors like housing income, poverty rates and population demographics, and uncover correlations and patterns that shed light on the broader societal context.

This will be achieved through the use of data visualizations to contribute to a deeper understanding of police shootings.


## Introduction
*Introduction: The introduction provides a clear explanation of the question and the dataset used to answer the question, including a description of all relevant variables in the dataset. (3 points)

Police shootings are a sad reality of policing practices that date back decades. While many are justified to protect the lives of innocent civilians, there has been an uptick in recent years of police shooting people whom they perceive to be a threat in the heat of the moment, but are proven innocent after the fact. These shootings can often be hard to analyze because there are many psychological factors that come into play when an officer decides to discharge their weapon, such as their training, the perceived threat, and an officer's internalized prejudice. Without further information on the psyche of a police officer, the reason why they discharge their weapon cannot be determined. However, their actions and surroundings can be analyzed to provide insight into how and why a shooting (unjustified or not) occurs.

The police shooting data set is sourced from a Kaggle dataset originally created by Karolina Wullum. The dataset consists of shootings from all fifty states and DC. 

This analysis aims to understand the dataset on police killings, extending its scope beyond the original focus tracked by the Washington Post. Socioeconomic factors like housing income, poverty rates, and population demographics will be used to understand and contextualize any patterns that appear. The goal is to explore how socioeconomic factors can play into incidents of police violence.

By incorporating these additional factors, correlations and patters can be uncovered that provide insight into the broader societal context surrounding police use of force. Through data visualization and analysis,  contributions to a deeper understanding of these issues and support evidence-based policy-making for promoting equity and justice.


(^above needs significant rewordin to fit the rubric requirements. Avoid 1st person...)

## Question 1

### Description of Analysis

### Plot 1
```{r}
#| label: Q1-plot1
#| message: false

```

### Plot 2a

```{r}
#| label: Q1-plot2a
#| message: false

```

### Plot 2b

```{r}
#| label: Q1-plot2b
#| message: false

```

### Discussion

## Question 2

### Description of Analysis

### Plot 1

```{r}
#| label: Q2-plot1
#| message: false

```

### Plot 2

```{r}
#| label: Q2-plot2
#| message: false

```

### Discussion