---
title: "Proposal title"
subtitle: "Proposal"
author: 
  - name: "RoboCops - Peter , Sachin , Arun , Surajit , Christian , Valérie"
    affiliations:
  - name: "School of Information, University of Arizona"
description: "Project description"
format:
  html:
    code-tools: true
    code-overflow: wrap
    code-line-numbers: true
    embed-resources: true
editor: visual
code-annotations: hover
execute:
  warning: false
---

```{r}
#| label: load-pkgs
#| message: false
#| echo: false
pacman::p_load(kableExtra)

```

## Introduction

The police shooting data set is sourced from a Kaggle dataset, originally created by Karolina Wullum. The data was originally logged by The Washington Post after the Black Lives Matter movement was born out of the shooting of Michael Brown in 2014. The data set contains information on fatal shootings that occurred from 2015 to 2016 and highlights information on the victims as well as the demographics of the city they were shot in. The goal of this project is to contextualize police shootings to further understand how and why these unfortunate situations occur.

## Why the Police Killings dataset?

Police shootings are a sad reality of policing practices that back decades. While many are justified to protect the lives of innocent civilians, there has been an uptick in recent years of police shooting people whom they perceive to be a threat in the heat of the moment, but are proven innocent after the fact. These shootings can often be hard to analyze because there are many psychological factors that come into play when an officer decides to discharge their weapon, such as their training, the perceived threat, and even an officer's internalized prejudice. Without further information on the psyche of a police officer, the reason why they discharge their weapon cannot be determined. However, their actions and surroundings can be analyzed to provide insight into how and why a shooting occurs.

## Analysis Goals

Our analysis aims to understand the dataset on police killings, extending its scope beyond the original focus tracked by the Washington Post. We plan to include socio-economic factors like housing income, poverty rates, and population demographics. While initially associated with the Black Lives Matter movement, our goal is to explore how socio-economic factors intersect with incidents of police violence.

By incorporating these additional factors, we hope to uncover correlations and patterns that provide insight into the broader societal context surrounding police use of force. Through data visualization and analysis, we aim to contribute to a deeper understanding of these issues and support evidence-based policy-making for promoting equity and justice

## Dataset

### Main Dataset

```{r}
#| label: load-dataset
#| message: false

# Load in the datasets
fatality <- read.csv("data/PoliceKillingsUS.csv", na.strings = "")
median_income <- read.csv("data/MedianHouseholdIncome2015.csv")
poverty_perc <- read.csv("data/PercentagePeopleBelowPovertyLevel.csv")
hs_perc <- read.csv("data/PercentOver25CompletedHighSchool.csv")
race_perc <- read.csv("data/ShareRaceByCity.csv")
state_pop <- read.csv("data/Population_Estimate_data_Statewise_2010-2023.csv")
```

The PoliceKilllingsUS.csv is a dataset sourced from Kaggle using the following link:

<a href=https://www.kaggle.com/datasets/kwullum/fatal-police-shootings-in-the-us/data>Fatal Police Shootings in the US</a>

The .csv contains data between 2015 and 2017 and was compiled by The Washington Post. The dataset is comprised of 14 variables and 2535 shootings. An in depth description of the variables can be seen in the following table:

```{r}
#| label: dataset-table
#| echo: false
#| message: false

# Create columns for data
fatality_var <- colnames(fatality)

fatality_desc <- c("Data ID",
                   "Name of deceased",
                   "Date of shooting",
                   "Means of death",
                   "Weapon/Tool",
                   "Age of deceased",
                   "Gender of deceased",
                   "Race of deceased",
                   "City",
                   "State",
                   "Sign of mental illness during incident",
                   "Deceased's threat level",
                   "Method to evade police",
                   "Reports of police with body camera")

fatality_type <- c("numeric",
                   "character",
                   "date",
                   "categorical",
                   "categorical",
                   "numeric",
                   "logical",
                   "categorical",
                   "categorical",
                   "categorical",
                   "logical",
                   "categorical",
                   "categorical",
                   "logical")

fatality_values <- c("2535 entries",
                     "Names",
                     "dd/mm/yy from 2015-2017",
                     "shot, shot and tasered",
                     "69 categories",
                     "6-91 years old",
                     "F/M",
                     "A, B, H, N, O, W, Unknown",
                     "1417 cities",
                     "State 2 letter abbreviations and DC",
                     "True/False",
                     "attack, other, undetermined",
                     "Car, Foot, Not fleeing, Other",
                     "True/False")

# Bind the vectors into a dataframe
fatality_table <- as.data.frame(cbind(fatality_var,
                                      fatality_desc,
                                      fatality_type,
                                     fatality_values))
colnames(fatality_table) <- c("Variable", "Description", "Data Type", "Values")
row.names(fatality_table) <- seq(1:14)

# Generate a nice table with kable
my_table <- kable(fatality_table, "html") %>%
  kable_styling(full_width = FALSE)

# Display the table
my_table

```

#### Data Cleaning/Wrangling

According to another Kaggle repository on the same topic and sourcing the same resources (<a href=https://www.kaggle.com/datasets/mrmorj/data-police-shootings>https://www.kaggle.com/datasets/mrmorj/data-police-shootings</a>), race codes unknown values as "". To avoid confusion, the values are recoded as Unknown using the `mutate` function.

Additionally, a secondary dataframe was generated by grouping the dataset by `city` and  `race` and acquiring the counts. This will allow us multiple avenues of analyses such as ascertaining whether or not the proportions of shootings by race in each city differs from the proportions of race in the cities in our census datasets. 

### City Census Datasets

In addition to the dataset containing the fatal police shooting, there are four additional 2015 census datasets each containing over 29000 observations:

-   MedianHouseholdIncome2015.csv

-   PercentagePeopleBelowPovertyLevel.csv

-   PercentOver25CompletedHighSchool.csv

-   ShareRaceByCity.csv

```{r}
#| label: census-table
#| echo: false
#| message: false

# Create columns for data
census_data <- c("PercentOver25CompletedHighSchool.csv",
                 "MedianHouseholdIncome2015.csv",
                 "PercentagePeopleBelowPovertyLevel.csv",
                 "ShareRaceByCity.csv")

census_desc <- c("Contains the percentage of individuals 25 years and older that are highschool graduates for cities",
                 "Contains the median income for cities",
                 "Contains the percentage of the population below poverty for cities",
                 "Contains the demographics for cities")

census_n <- c(nrow(hs_perc),
              nrow(median_income),
              nrow(poverty_perc),
              nrow(race_perc))

# Bind the vectors into a dataframe
census_table <- as.data.frame(cbind(census_data, census_desc, census_n))
colnames(census_table) <- c("Data", "Description", "Number of Cities")
row.names(census_table) <- seq(1:4)

# Generate a nice table with kable
my_table <- kable(census_table, "html") %>%
  kable_styling(full_width = FALSE)

# Display the table
my_table

```

Each census dataset contains two common columns: `Geographic.Area` and `City`. The values for `City` is name of the city and a classifier (ie Alexander City city).

In addition to these two columns, each dataset contains information regarding the city's demographic information: percent of highschool graduates, median income, poverty rate, and proportion of races (white, black, native american, asian, hispanic). Each of these values are numeric.

#### Data Cleaning/Wrangling

For ease of analysis, the four census datasets are concatenated together by city into one merged dataset.

Additionally, the `City` value is edited to remove the classifier from each label so that the city value between the main dataset and the census datasets will be equivalent. Doing so will ease analysis.

### State Census Data

In addition to city level census data, a statewise census dataset was collected. Population_Estimate_data_Statewise_2010-2023.csv contains population estimates from 2010-2023 for all fifty states, the District of Colombia, and Puerto Rico.

## Questions

The two questions you want to answer.

1.  How do socioeconomic factors such as median income, poverty rates, and educational relate with incidents of police use of force across different geographic areas?

2.  To what extent do racial demographics, including the proportions of White, Black, Hispanic, Asian, and Native American residents, correlate with incidents of police use of force within communities?

## Analysis plan

-   Question 1:

    -   By answering how poverty levels relate w.r.t policing killing in each state

    -   We overlay this w.r.t police instrument usage

    -   Sub-question 1: How does the percentage of people below the poverty line in each city correlate with the frequency and severity of police use of force incidents?

    -   Sub-question 2: Are there notable differences in the threat level perceived by police officers when encountering individuals from different socio-economic backgrounds, as indicated by poverty rates and educational attainment levels?

    -   Plot 1: Regression x = percent HS/ median income, y  = count of police killings grouped by city and race, can overlay lines with race

    -   Plot 2: half-eye plot (need more variables for regression)

    -   Plot 3: Bar Plot Comparing Socioeconomic Indicators

-   Question 2:

    -   Sub-question 1: Do certain racial demographics face disproportionately higher incidents of police use of force within specific cities, and if so, how does this correlate with socio-economic factors such as median household income and poverty rates?

    -   Sub-question 2: Are there discernible patterns in the manner of death (e.g., whether the deceased was shot or shot and tasered) based on racial demographics and geographical locations of police shootings?

    -   Plot 1: Leaflet plot, how police shooting have changed in five years

    -   Plot 2: Leaflet plot of police shooting w.r.t. Population of each state { Shooting Intensity = #Policeshootings/Population }

    -   Plot 3: focus on certain cities/states and overlay

    -   Plot 4: Heatmap of Police Use of Force Incidents by Racial Demographics

## Timeline

```{r}
#| label: timeline-table
#| echo: false
#| message: false

# Create vectors to represent the tasks
tasks <- c("Proposal-For Peer Review",
           "Peer Review",
           "Proposal-Updated",
           "Proposal-Final")

# Create vector to represent status
status <- c("",
            "",
            "",
            "")

# Create vector to represent ownership
member <- c("All",
            "All",
            "All",
            "All")

# Create vector to represent the deadline
due <- c("April 1",
         "April 1",
         "April 8",
         "April 15")

# Create vector to represent priority
priority <- c("High",
              "High",
              "Medium",
              "Low")

# Create vector to represent description
description <- c("Write proposal for peer review",
                 "Write up peer reviews for other groups",
                 "Update proposal using peer feedback",
                 "Update proposal using instructor feedback")

# Bind the columns
table_2 <- as.table(cbind(tasks, status, member, due, priority, description))
colnames(table_2) <- c("Task Name", "Status", "Assignee",
                       "Due", "Priority", "Summary")
row.names(table_2) <- seq(1:nrow(table_2))

# Generate a nice table with kable
timeline_table <- kable(table_2, "html") %>%
  kable_styling(full_width = FALSE)

# Display the table
timeline_table
```
