---
title: "Proposal title"
subtitle: "Proposal"
author: 
  - name: "RoboCops - Peter Yeh, Sachin Pandurang Patil, Arun Koundinya Parasa, Surajit Pal, Christian Ortmann, Valérie Naa Dei Okine"
    affiliations:
      - name: "School of Information, University of Arizona"
description: "Project description"
format:
  html:
    code-tools: true
    code-overflow: wrap
    code-line-numbers: true
    embed-resources: true
editor: visual
code-annotations: hover
execute:
  warning: false
---

```{r}
#| label: load-pkgs
#| message: false
#| echo: false
pacman::p_load(kableExtra)

```

## Introduction


## Why the Police Killings dataset?


## Analysis Goals


## Dataset
### Main Dataset

```{r}
#| label: load-dataset
#| message: false

# Load in the datasets
fatality <- read.csv("data/PoliceKillingsUS.csv", na.strings = "")
median_income <- read.csv("data/MedianHouseholdIncome2015.csv")
poverty_perc <- read.csv("data/PercentagePeopleBelowPovertyLevel.csv")
hs_perc <- read.csv("data/PercentOver25CompletedHighSchool.csv")
race_perc <- read.csv("data/ShareRaceByCity.csv")

```

The PoliceKilllingsUS.csv is a dataset sourced from Kaggle using the following
link:

<a href=https://www.kaggle.com/datasets/kwullum/fatal-police-shootings-in-the-us/data>Fatal Police Shootings in the US</a>

The .csv contains data between 2015 and 2017 and was compiled by The Washington
Post. The dataset is comprised of 14 variables and 2535 shootings. An in depth
description of the variables can be seen in the following table:

```{r}
#| label: dataset-table
#| echo: false
#| message: false

# Create columns for data
fatality_var <- colnames(fatality)

fatality_desc <- c("Data ID",
                   "Name of deceased",
                   "Date of shooting",
                   "Means of death",
                   "Weapon/Tool",
                   "Age of deceased",
                   "Gender of deceased",
                   "Race of deceased",
                   "City",
                   "State",
                   "Sign of mental illness during incident",
                   "Deceased's threat level",
                   "Method to evade police",
                   "Reports of police with body camera")

fatality_type <- c("numeric",
                   "character",
                   "date",
                   "categorical",
                   "categorical",
                   "numeric",
                   "logical",
                   "categorical",
                   "categorical",
                   "categorical",
                   "logical",
                   "categorical",
                   "categorical",
                   "logical")

fatality_values <- c("2535 entries",
                     "Names",
                     "dd/mm/yy from 2015-2017",
                     "shot, shot and tasered",
                     "69 categories",
                     "6-91 years old",
                     "F/M",
                     "A, B, H, N, O, W, None",
                     "1417 cities",
                     "State 2 letter abbreviations and DC",
                     "True/False",
                     "attack, other, undetermined",
                     "Car, Foot, Not fleeing, Other",
                     "True/False")

# Bind the vectors into a dataframe
fatality_table <- as.data.frame(cbind(fatality_var,
                                      fatality_desc,
                                      fatality_type,
                                     fatality_values))
colnames(fatality_table) <- c("Variable", "Description", "Data Type", "Values")
row.names(fatality_table) <- seq(1:14)

# Generate a nice table with kable
my_table <- kable(fatality_table, "html") %>%
  kable_styling(full_width = FALSE)

# Display the table
my_table

```

#### Data Cleaning/Wrangling

### Census Dataset
In addition to the dataset containing the fatal police shooting, there are four
additional 2015 census datasets each containing over 29000 observations:
MedianHouseholdIncome2015.csv, PercentagePeopleBelowPovertyLevel.csv,
PercentOver25CompletedHighSchool.csv, and ShareRaceByCity.csv.

```{r}
#| label: census-table
#| echo: false
#| message: false

# Create columns for data
census_data <- c("PercentOver25CompletedHighSchool.csv",
                 "MedianHouseholdIncome2015.csv",
                 "PercentagePeopleBelowPovertyLevel.csv",
                 "ShareRaceByCity.csv")

census_desc <- c("Contains the percentage of highschool graduates for cities",
                 "Contains the median income for cities",
                 "Contains the percentage of the population below poverty for cities",
                 "Contains the demographics for cities")

census_n <- c(nrow(hs_perc),
              nrow(median_income),
              nrow(poverty_perc),
              nrow(race_perc))

# Bind the vectors into a dataframe
census_table <- as.data.frame(cbind(census_data, census_desc, census_n))
colnames(census_table) <- c("Data", "Description", "Number of Cities")
row.names(census_table) <- seq(1:4)

# Generate a nice table with kable
my_table <- kable(census_table, "html") %>%
  kable_styling(full_width = FALSE)

# Display the table
my_table

```

Each census dataset contains two common columns: `Geographic.Area` and `City`.
City` is coded in such a way that the column contains the name of the city and a
classifier (city, town, CDP, etc.).

In addition to these two columns, each dataset contains information regarding
the city's demographic information: percent of highschool graduates,
median income, poverty rate, and proportion of races
(white, black, native american, asian, hispanic). Each of these values are
numeric.

#### Data Cleaning/Wrangling

## Questions

The two questions you want to answer.

## Analysis plan

-   A plan for answering each of the questions including the variables involved, variables to be created (if any), external data to be merged in (if any).

## Timeline

```{r}
#| label: timeline-table
#| echo: false
#| message: false

# Create vectors to represent the tasks
tasks <- c("Proposal-For Peer Review",
           "Peer Review",
           "Proposal-Updated",
           "Proposal-Final")

# Create vector to represent status
status <- c("",
            "",
            "",
            "")

# Create vector to represent ownership
member <- c("All",
            "All",
            "All",
            "All")

# Create vector to represent the deadline
due <- c("April 1",
         "April 1",
         "April 8",
         "April 15")

# Create vector to represent priority
priority <- c("High",
              "High",
              "Medium",
              "Low")

# Create vector to represent description
description <- c("Write proposal for peer review",
                 "Write up peer reviews for other groups",
                 "Update proposal using peer feedback",
                 "Update proposal using instructor feedback")

# Bind the columns
table_2 <- as.table(cbind(tasks, status, member, due, priority, description))
colnames(table_2) <- c("Task Name", "Status", "Assignee",
                       "Due", "Priority", "Summary")
row.names(table_2) <- seq(1:nrow(table_2))

# Generate a nice table with kable
timeline_table <- kable(table_2, "html") %>%
  kable_styling(full_width = FALSE)

# Display the table
timeline_table
```
